Extended Data

### 2022-05-22 doc/manuscript/Dow_Extended_Data.Rmd
italicize variables & avoid latex
-Italic font used for variables, genes and non-English languages; bold font for vectors
- Equations and special characters in Word in ‘normal text’ or Symbol font (Word Equation Editor avoided where possible)

```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring-porous","diffuse-porous","ring-porous", "diffuse-porous"), stringsAsFactors = FALSE, check.names = FALSE)
kable(mean_growth_parameters, caption = mean_growth_parameters_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
    column_spec(1, width = "4cm") %>%
    pack_rows("Stem Growth", 2, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Temperature sensitivity* (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
    add_indent(13:16)%>%
    add_indent(18:21)%>%
kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (CTW, Fig. 1). Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward. We used the CTW of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.

@@

![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the CTW. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

@@

![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=100% }

@@

![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Figure presents the same results as main manuscript Fig. 3 but extends back to include the previous year. Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=100% }

>>>>

```{r mean_growth_parameters_table,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
mean_growth_parameters <- read.csv("tables_figures/phenology_parameters.csv",  col.names = c("", "ring-porous","diffuse-porous","ring-porous", "diffuse-porous"), stringsAsFactors = FALSE, check.names = FALSE)
kable(mean_growth_parameters, caption = mean_growth_parameters_caption ,  format = "latex", booktabs = TRUE,  escape = F) %>%
add_header_above(c(" ",  "SCBI" = 2, "Harvard Forest"=2 )) %>%
    column_spec(1, width = "4cm") %>%
    pack_rows("Stem Growth", 2, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Canopy Foliage (ecosystem level)", 8, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
    pack_rows("Temperature sensitivity* (days/°C)", 12, 20, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
    add_indent(13:16)%>%
    add_indent(18:21)%>%
kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
```
$^\ast$Temperature sensitivity refers to the sensitivity of a parameter to the mean maximum temperature ($T_{max}$) during the critical temperature window (_CTW_, Fig. 1). Because leaf phenology measurements were derived from satellite imagery and included both ring- and diffuse-porous trees, the selection of a critical window was less straightforward. We used the _CTW_ of the porosity group containing the dominant canopy species at each site: *Liriodendron tulipifera* (diffuse porous) at SCBI and *Quercus rubra* (ring porous) at Harvard Forest.

@@

![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

@@

![**Extended Data Figure 6 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean minimum temperatures ($T_{min}$), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the bootstrapped correlation between monthly $T_{min}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April $T_{min}$. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Species analyzed and numbers of significant correlations to $T_{min}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmin.png){ width=100% }

@@

![**Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures ($T_{max}$) of the current and past year, for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America** (`r EDF_map `). Colors indicate the bootstrapped correlation between monthly $T_{max}$ and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April $T_{max}$. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF) (`r EDT_species`). Figure presents the same results as main manuscript Fig. 3 but extends back to include the previous year. Species analyzed and numbers of significant correlations to $T_{max}$ are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax_extended.png){ width=100% }
